Exploring dimensions of biodiversity in Japanese ferns

### 2020-07-14 presentations/bsa_2020-07-27/index.Rmd
Update slides

```{r load-packages, cache = FALSE}
library(conflicted)
library(magick)
library(ggtext)
library(drake)
library(knitr)
library(patchwork)
library(scales)
library(scico)
library(magrittr)
library(broom)
library(assertr)
library(sf)
library(rcartocolor)
library(tidyverse)

@@

.pull-left-wide[
# Exploring dimensions of biodiversity in Japanese ferns
<br>
## Joel H. Nitta<sup>1</sup>, Atsushi Ebihara<sup>2</sup>

<sup>1</sup>University of Tokyo<br>
<span style = 'font-size: 80%;'>https://joelnitta.com</span>

<sup>2</sup>National Museum of Nature and Science, Japan

Botanical Society of America<br><span style = 'font-size: 80%;'>2020.07.27</span>
]

@@

## Biodiversity is a two-sided coin

Characterizing the **distribution of biodiversity** is needed to:

- understand the **processes** generating it

- understand how to **conserve** it

## Most studies of biodiversity count species

However, taxonomic richness is only one dimension of biodiversity!

@@

This requires comprehensive sampling of taxa, DNA, and traits

@@

## Ferns of Japan

@@

@@

- Extensive study by both scientists and amateur societies

@@

## Ferns of Japan

.pull-left[
- One of the best-collected fern floras in the world

- Extensive study by both scientists and amateur societies
]

.pull-right[
```{r ja-pterido-coll-curve}
# Format fern abundances (no hybrids) as numeric vector
fern_abun <-
  occ_point_data_ferns %>%
  group_by(taxon) %>%
  count(sort = TRUE) %>%
  pull(n)

# Run iNEXT
inext_out <- iNEXT::iNEXT(fern_abun, q=0, datatype="abundance")

# create subset of observed points to add to plot as dots
observed_data <- inext_out$iNextEst[inext_out$iNextEst$method == "observed",]

# check estimated completeness
est_completeness <- inext_out$iNextEst %>% filter(method == "observed") %>% pull(SC) %>% scales::percent()
rich_interval <- glue::glue("{observed_data$qD.LCL %>% round} to {observed_data$qD.UCL %>% round}")
completeness_interval <- glue::glue("{observed_data$SC.LCL %>% scales::percent()} to {observed_data$SC.UCL %>% scales::percent()}")

# Make iNEXT plot
inext_out$iNextEst %>%
  filter(method != "observed") %>%
  ggplot(aes(x=m, y=qD)) +
  geom_ribbon(aes(ymin = qD.LCL, ymax = qD.UCL), fill = "grey70") +
  geom_line(aes(linetype=method), size=0.8) +
  scale_linetype_manual(values=c("dotted", "solid", "solid")) +
  geom_point(data=observed_data, size=3) +
  scale_y_continuous("Richness") +
  scale_x_continuous("No. specimens") +
  labs(
    title = "Species collection curve",
    subtitle = "Japanese ferns (no hybrids)"
  ) +
  ggplot2::theme_bw(base_size = 24) +
  theme(
    panel.grid.minor = ggplot2::element_blank(),
    axis.text.x = ggplot2::element_text(colour="black"),
    axis.text.y = ggplot2::element_text(colour="black"),
    legend.position="bottom", 
    legend.title=element_blank()
    )

```
]

@@

@@

```{r lat-bins}
# Bin SES diversity results by latitudinal band (0.2 degree width)
ses_div_binned <-
  ses_div_ferns_spatial %>%
  mutate(
    longitude = sf::st_centroid(geometry) %>% st_coordinates %>% magrittr::extract(,1),
    latitude = sf::st_centroid(geometry) %>% st_coordinates %>% magrittr::extract(,2)) %>%
  as_tibble %>%
  dplyr::select(-geometry) %>%
  mutate(
    latitude = cut_width(latitude, width = 0.2, closed = "left") %>%
      str_split(",") %>%
      map_chr(1) %>%
      parse_number) %>%
  group_by(latitude) %>%
  summarize(
    across(c(richness, pe_obs_rank, contains("obs_z")), ~mean(.x, na.rm = TRUE))
  )

# Get min and max latitude for setting common limits between
# maps and trendlines
ja_lat_min <- ses_div_binned %>% pull(latitude) %>% min()
ja_lat_max <- ses_div_binned %>% pull(latitude) %>% max()
```

@@

# - line tracking mean richness
b <-
  ggplot(ses_div_binned, aes(y = latitude, x = richness)) +
  geom_path() +
  labs(
    x = "No. taxa"
  ) +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_line(size = 0.1, color = "dark grey"),
    panel.grid.minor = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_blank()
  ) +
  ylim(ja_lat_min, ja_lat_max)

@@

---
## Other raw values are correlated with richness

```{r raw-div, out.width = "120%", fig.asp = 0.7}
# Richness
a <- ggplot(ses_div_ferns_spatial, aes(fill = richness)) +
  geom_sf(color = "transparent") +
  scale_fill_scico(palette = "lajolla") + 
  labs(title = "Taxonomic richness") +
  simple_map_theme()

b <- ggplot(ses_div_ferns_spatial, aes(fill = pd_obs)) +
  geom_sf(color = "transparent") +
  scale_fill_scico(palette = "lajolla") + 
  labs(title = "Phylogenetic diversity") +
  simple_map_theme()

c <- ggplot(ses_div_ferns_spatial, aes(fill = fd_obs)) +
  geom_sf(color = "transparent") +
  scale_fill_scico(palette = "lajolla") + 
  labs(title = "Functional diversity") +
  simple_map_theme()

@@

a + b + c +
  grid::textGrob("...as expected") + d + e +
  plot_layout(nrow = 2, heights = c(0.5, 0.5))
```

.pull-left[
$SES = \frac{x - mean(x)}{sd(x)}$
]

.pull-right[
```{r ses-example}
null <- tibble(value = rnorm(10000))

obs_1 <- 0.95
obs_2 <- 0.5
obs_3 <- 0.05

ggplot(null, aes(x = value)) +
  geom_histogram(bins = 50) +
  geom_vline(xintercept = 0)
```
]

@@

# - line tracking mean SES RPD
b <-
  ggplot(ses_div_binned, aes(y = latitude, x = pd_obs_z)) +
  geom_path() +
  geom_vline(xintercept = 0, linetype="dashed") +
  labs(
    x = "SES PD"
  ) +
  ylim(24, 46) +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_line(size = 0.1, color = "dark grey"),
    panel.grid.minor = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_blank()
  )

@@

@@

# - line tracking mean SES RPD
b <-
  ggplot(ses_div_binned, aes(y = latitude, x = rpd_obs_z)) +
  geom_path() +
  geom_vline(xintercept = 0, linetype="dashed") +
  labs(
    x = "SES RPD"
  ) +
  ylim(24, 46) +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_line(size = 0.1, color = "dark grey"),
    panel.grid.minor = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_blank()
  )

@@

---
## <u>**R**</u>elative <u>**F**</u>unctional <u>**D**</u>iversity: functional analog of RPD

RFD also high in the south

@@

# - line tracking mean SES RPD
b <-
  ggplot(ses_div_binned, aes(y = latitude, x = rfd_obs_z)) +
  geom_path() +
  geom_vline(xintercept = 0, linetype="dashed") +
  labs(
    x = "SES RFD"
  ) +
  ylim(24, 46) +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_line(size = 0.1, color = "dark grey"),
    panel.grid.minor = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_blank(),
    plot.margin = margin(l = -1, unit = "inch")
  )

@@

---
## Phylogenetic endemism: PD weighted by range size

@@

```{r pe, out.width = "120%"}
# PE
# - map
a <- ggplot(ses_div_ferns_spatial, aes(fill = pe_obs_rank)) +
  geom_sf(color = "transparent") +
  scale_fill_scico(palette = "lajolla") + 
  labs(fill = "PE Rank") +
  scale_y_continuous(labels = scales::label_number(suffix = "\u00b0N", accuracy = 1)) +
  scale_x_continuous(labels = scales::label_number(suffix = "\u00b0E", accuracy = 1)) +
  ylim(ja_lat_min, ja_lat_max) +
  map_theme_pp() +
  theme(
    legend.title = element_text(size = 20/.pt),
    legend.text = element_text(size = 16/.pt),
    legend.justification=c(0,1), legend.position=c(0,1)
  )

# - line tracking mean richness
b <-
  ggplot(ses_div_binned, aes(y = latitude, x = pe_obs_rank)) +
  geom_path() +
  labs(
    x = "PE Rank"
  ) +
  ylim(ja_lat_min, ja_lat_max) +
  trend_line_theme()

@@

---
## Bioregions

Most sites fall into four bioregions

```{r tax-bioregion, out.width = "120%", fig.asp = 0.7}
ggplot(ses_div_ferns_spatial, aes(fill = taxonomic_cluster)) +
  geom_sf(color = "transparent") +
  scale_fill_carto_d(name = "Cluster", type = "qualitative", palette = "Bold") +
  scale_y_continuous(labels = scales::label_number(suffix = "\u00b0N", accuracy = 1)) +
  scale_x_continuous(labels = scales::label_number(suffix = "\u00b0E", accuracy = 1)) +
  map_theme_light()
```

@@

```{r protected-areas, out.width = "120%", fig.asp = 0.7}
japan_box <- sf::st_bbox(ses_div_ferns_spatial)

@@

---

## Hotspots of taxonomic richness are well-protected

@@

# final plot: add inset
cowplot::ggdraw(richness_initial_plot) +
  cowplot::draw_plot(inset, 0.8, 0.45, .30, .30, hjust = 1, vjust = 1) 
```
---

## Hotspots of phylogenetic endemism are also well-protected

@@

>>>>

```{r load-packages-funcs, cache = FALSE}
library(conflicted)
library(magick)
library(ggtext)
library(drake)
library(knitr)
library(patchwork)
library(scales)
library(scico)
library(magrittr)
library(broom)
library(assertr)
library(sf)
library(rcartocolor)
library(tidyverse)

@@

.pull-left-wide[
# Exploring dimensions of biodiversity in Japanese ferns
<br>
## Joel H. Nitta<sup>1</sup>, Brent D. Misher<sup>2</sup>, Wataru Iwasaki<sup>1</sup>, Atsushi Ebihara<sup>3</sup>

<sup>1</sup>University of Tokyo, Japan<br>

<sup>2</sup>University of California, Berkeley, USA<br>

<sup>3</sup>National Museum of Nature and Science, Japan

<span style = 'font-size: 80%;'>https://joelnitta.com</span>
]

@@

## Biodiversity: more than just counting species

Species richness is only one dimension of biodiversity!

@@

<img src="images/not_just_species.png" alt="drawing" width="800"/>

@@

## Biodiversity: more than just counting species

Species richness is only one dimension of biodiversity!

**This requires comprehensive sampling of taxa, DNA, and traits**

## Ferns of Japan: an ideal system for studying biodiversity

@@

- Diverse flora (ca. 675 spp.)

@@

- Data available for DNA, distribution, and traits

@@

## (Very brief) introduction to the biogeography of Japan

- Wide variation in climate from north (subarctic) to south (subtropical)

- Wide variation in elevation

- Tertiary relict flora (never glaciated)

- Main islands continental, southern islands oceanic

@@

---
class: big-margin
## Data sources

```{r summarize-traits}
trait_summary <- make_trait_summary(traits_for_dist)

n_binary <- trait_summary %>% filter(trait_type == "binary") %>% nrow()
n_cont <- trait_summary %>% filter(trait_type == "continuous") %>% nrow()
n_qual <- trait_summary %>% filter(trait_type == "qualitative") %>% nrow()
```

**Occurrence data**

- Specimens at herbaria in Japan

- `r nrow(occ_point_data_ferns) %>% scales::number(big.mark = ",")` specimens (`r occ_point_data_ferns %>% pull(taxon) %>% n_distinct()` taxa)

- Convert to 0.2° (ca. 20 km x 20 km) presence-absence matrix

---
class: big-margin
## Data sources

**Phylogenetic tree**

- *rbcL* sequences from Japanese specimens (Ebihara and Nitta 2019)

- Combine with broad sampling from genbank (ca. 60 genes x 4,000 taxa) (Nitta et al., in prep.)

- Infer phylogeny with IQ-TREE (Nguyen et al. 2015)

- Date with 26 fossil calibration points (Testo and Sundue 2016) using treePL (Smith and O'Meara 2012)

- Subset to only Japanese taxa

## Data sources

**Traits**

- `r nrow(trait_summary)` morphological traits mostly used for ID (Ebihara and Nitta 2019)

  - `r n_binary` binary (e.g., false indusium)

  - `r n_cont` continuous (e.g., stipe length)

  - `r n_qual` qualitative (e.g., indusium shape, lamina texture)

---
class: middle, inverse, big-margin

# How is biodiversity distributed?

@@

```{r centroids}
# Convert diversity metrics dataframe to centroids for plotting trends
ses_div_centroids <-
  ses_div_ferns_spatial %>%
  mutate(
    longitude = sf::st_centroid(geometry) %>% st_coordinates %>% magrittr::extract(,1),
    latitude = sf::st_centroid(geometry) %>% st_coordinates %>% magrittr::extract(,2)) %>%
  as_tibble %>%
  dplyr::select(-geometry)

# Get min and max latitude for setting common limits between
# maps and trendlines
ja_lat_min <- ses_div_centroids %>% pull(latitude) %>% min()
ja_lat_max <- ses_div_centroids %>% pull(latitude) %>% max()
```

@@

# - latitudinal trends
b <- ggplot(ses_div_centroids, aes(x = latitude, y = richness)) +
  geom_point(color = "grey50", shape = 1 ) +
  geom_smooth(se = FALSE) +
  xlim(ja_lat_min, ja_lat_max) +
  coord_flip() +
  labs(y = "No. taxa") +
  trend_line_theme()

@@

.footnote[Trendline: General additive model (GAM) with cubic splines]
---
## Other raw values are correlated with richness

```{r raw-div-1, out.width = "120%", fig.asp = 0.7}
# Richness
a <- ggplot(ses_div_ferns_spatial, aes(fill = richness)) +
  geom_sf(color = "transparent") +
  scale_fill_scico(palette = "lajolla") + 
  labs(title = "Taxonomic richness") +
  simple_map_theme()

b <- ggplot(ses_div_ferns_spatial, aes(fill = pd_obs)) +
  geom_sf(color = "transparent") +
  scale_fill_scico(palette = "lajolla") + 
  labs(
    title = "Phylogenetic diversity",
    subtitle = "Faith's PD") +
  simple_map_theme()

c <- ggplot(ses_div_ferns_spatial, aes(fill = fd_obs)) +
  geom_sf(color = "transparent") +
  scale_fill_scico(palette = "lajolla") + 
  labs(
    title = "Functional diversity",
    subtitle = "Faith's PD measured\non trait dendrogram") +
  simple_map_theme()

a + b + c +
  plot_spacer() + plot_spacer() + plot_spacer() +
  plot_layout(nrow = 2, heights = c(0.5, 0.5))
```

---
## Other raw values are correlated with richness... as expected

```{r raw-div-2, out.width = "120%", fig.asp = 0.7}
# Richness
a <- ggplot(ses_div_ferns_spatial, aes(fill = richness)) +
  geom_sf(color = "transparent") +
  scale_fill_scico(palette = "lajolla") + 
  labs(title = "Taxonomic richness") +
  simple_map_theme()

b <- ggplot(ses_div_ferns_spatial, aes(fill = pd_obs)) +
  geom_sf(color = "transparent") +
  scale_fill_scico(palette = "lajolla") + 
  labs(
    title = "Phylogenetic diversity",
    subtitle = "Faith's PD") +
  simple_map_theme()

c <- ggplot(ses_div_ferns_spatial, aes(fill = fd_obs)) +
  geom_sf(color = "transparent") +
  scale_fill_scico(palette = "lajolla") + 
  labs(
    title = "Functional diversity",
    subtitle = "Faith's PD measured\non trait dendrogram") +
  simple_map_theme()

@@

a + b + c +
  plot_spacer() + d + e +
  plot_layout(nrow = 2, heights = c(0.5, 0.5))
```

.footnote[Trendline: GAM with log-transform]
---
class: big-margin
## Compare raw values to random expectation with<br> <u>**S**</u>tandard <u>**E**</u>ffect <u>**S**</u>ize

.large[
.center[
$SES = \frac{obs - mean(null)}{sd(null)}$
]
]

SES = 0: not different from random

SES > 0: more diverse than expected

SES < 0: less diverse than expected

Null distribution: 999 communities randomized using "Independent Swap" (Gotelli, 2000)

@@

# - latitudinal trends
b <- ggplot(ses_div_centroids, aes(x = latitude, y = pd_obs_z)) +
  geom_point(color = "grey50", shape = 1 ) +
  geom_hline(yintercept = 0, linetype="dashed") +
  geom_smooth(se = FALSE) +
  xlim(ja_lat_min, ja_lat_max) +
  coord_flip() +
  labs(y = "SES PD") +
  trend_line_theme()

@@

.footnote[Trendline: GAM with cubic splines]
---
## Compare branch lengths with <u>**R**</u>elative <u>**P**</u>hylogenetic <u>**D**</u>iversity

@@

# - latitudinal trends
b <- ggplot(ses_div_centroids, aes(x = latitude, y = rpd_obs_z)) +
  geom_point(color = "grey50", shape = 1 ) +
  geom_hline(yintercept = 0, linetype="dashed") +
  geom_smooth(se = FALSE) +
  xlim(ja_lat_min, ja_lat_max) +
  coord_flip() +
  labs(y = "SES RPD") +
  trend_line_theme()

@@

.footnote[Trendline: GAM with cubic splines]
---
## <u>**R**</u>elative <u>**F**</u>unctional <u>**D**</u>iversity: functional analog of RPD

Southern islands have high RFD

@@

# - latitudinal trends
b <- ggplot(ses_div_centroids, aes(x = latitude, y = rfd_obs_z)) +
  geom_point(color = "grey50", shape = 1 ) +
  geom_hline(yintercept = 0, linetype="dashed") +
  geom_smooth(se = FALSE) +
  xlim(ja_lat_min, ja_lat_max) +
  coord_flip() +
  labs(y = "SES RFD") +
  trend_line_theme()

@@

.footnote[Trendline: GAM with cubic splines]
---
## Phylogenetic endemism: PD weighted by range size

@@

```{r pe, out.width = "120%"}
# PE
# - map
a <- ggplot(ses_div_ferns_spatial, aes(fill = pe_obs_p)) +
  geom_sf(color = "transparent") +
  scale_fill_scico(palette = "lajolla", labels = scales::percent) + 
  labs(fill = "PE percentile") +
  scale_y_continuous(labels = scales::label_number(suffix = "\u00b0N", accuracy = 1)) +
  scale_x_continuous(labels = scales::label_number(suffix = "\u00b0E", accuracy = 1)) +
  ylim(ja_lat_min, ja_lat_max) +
  map_theme_pp() +
  theme(
    legend.title = element_text(size = 20/.pt),
    legend.text = element_text(size = 16/.pt),
    legend.justification=c(0,1), legend.position=c(0,1)
  )

# - line tracking mean richness
b <- ggplot(ses_div_centroids, aes(x = latitude, y = pe_obs_p)) +
  geom_point(color = "grey50", shape = 1 ) +
  geom_hline(yintercept = 0.95, linetype="dashed") +
  geom_smooth(se = FALSE) +
  annotate("label", y = 0.95, x = ja_lat_max, label = "95%", size = 2) + 
  xlim(ja_lat_min, ja_lat_max) +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "PE percentile") +
  coord_flip() +
  trend_line_theme()

@@

.footnote[Trendline: GAM with cubic splines]
---
class: middle, inverse, big-margin

# How is biodiversity structured?

---
## Most sites fall into four bioregions

```{r tax-bioregion, out.width = "120%", fig.asp = 0.7}
ggplot(ses_div_ferns_spatial, aes(fill = taxonomic_cluster)) +
  geom_sf(color = "transparent") +
  scale_fill_carto_d(name = "Cluster", type = "qualitative", palette = "Bold") +
  scale_y_continuous(labels = scales::label_number(suffix = "\u00b0N", accuracy = 1)) +
  scale_x_continuous(labels = scales::label_number(suffix = "\u00b0E", accuracy = 1)) +
  map_theme_pp_light() +
  labs(
    title = "Taxonomic bioregions"
  )
```

---
## Taxonomic and phylogenetic bioregions are similar

```{r compare-bioregion, out.width = "120%", fig.asp = 0.7}
a <- ggplot(ses_div_ferns_spatial, aes(fill = taxonomic_cluster)) +
  geom_sf(color = "transparent") +
  scale_fill_carto_d(name = "Cluster", type = "qualitative", palette = "Bold") +
  scale_y_continuous(labels = scales::label_number(suffix = "\u00b0N", accuracy = 1)) +
  scale_x_continuous(labels = scales::label_number(suffix = "\u00b0E", accuracy = 1)) +
  map_theme_pp_light() +
  labs(
    title = "Taxonomic bioregions"
  ) +
  theme(legend.position = "none")

# (If needed)
# Swap first two colors to get them to match between taxonomic and phylogenetic plots
# cols_old <- rcartocolor::carto_pal(n = 8,  "Bold")
# cols <- cols_old
# cols[1] <- cols_old[2]
# cols[2] <- cols_old[1]

b <- ggplot(ses_div_ferns_spatial, aes(fill = phylo_cluster)) +
  geom_sf(color = "transparent") +
  # scale_fill_manual(values = cols) +
  scale_fill_carto_d(name = "Cluster", type = "qualitative", palette = "Bold") +
  scale_y_continuous(labels = scales::label_number(suffix = "\u00b0N", accuracy = 1)) +
  scale_x_continuous(labels = scales::label_number(suffix = "\u00b0E", accuracy = 1)) +
  labs(title = "Phylogenetic bioregions") +
  map_theme_pp_light() +
  theme(
    axis.text.y = element_blank(),
    legend.position = "none")

a + b 
```

---
## Bioregions mostly match traditionally-defined forest types

.pull-left[
Taxonomic bioregions

```{r tax-bioregion-forest}
ggplot(ses_div_ferns_spatial, aes(fill = taxonomic_cluster)) +
  geom_sf(color = "transparent") +
  scale_fill_carto_d(name = "Cluster", type = "qualitative", palette = "Bold") +
  scale_y_continuous(labels = scales::label_number(suffix = "\u00b0N", accuracy = 1)) +
  scale_x_continuous(labels = scales::label_number(suffix = "\u00b0E", accuracy = 1)) +
  map_theme_pp_light() +
  theme(legend.position = "none")
```
]

.pull-right[
Forest types

<img src="images/forest_types_eng.png" alt="drawing" width="280"/>
]

.footnote[Shimizu 2014]

---
class: middle, inverse, big-margin

# How well is biodiversity protected?

@@

```{r protected-areas, out.width = "120%", fig.asp = 0.7, fig.retina=200}
japan_box <- sf::st_bbox(ses_div_ferns_spatial)

@@

---
## Hotspots of taxonomic richness are well-protected

@@

# final plot: add inset
cowplot::ggdraw(richness_initial_plot) +
  cowplot::draw_plot(inset, 0.8, 0.45, .30, .30, hjust = 1, vjust = 1) 
```

---
## Hotspots of phylogenetic endemism are also well-protected

@@

## Conclusions

How is biodiversity distributed?

- Taxonomic diversity highest in Kyushu

- Phylogenetic and functional diversity, phylogenetic endemism highest in Okinawa

How is biodiversity structured?

- Into four major clusters that track major forest types

How well is biodiversity protected?

- Well-protected in terms of status

- May be in danger from herbivory by deer

## Future directions

- Understand the role that hybridization has on diversification (ca. 400 described hybrids!)

- Clarify the role of fern gametophytes in community assembly at broad scales
